\Title

### 2023-06-01 Grimaldi-etal-BiolCybernetics.tex
Update Grimaldi-etal-BiolCybernetics.tex
intro with https://www.deepl.com/write

\keywords{\Keywords}
%
\maketitle
%
\section{Introduction}
%%%-----------------------------------------------------------------
\label{sec:intro}
%
The human brain has the remarkable property of being able to react efficiently, at any time while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference with artificial neural networks is the representation both use. Artificial convolutional neural networks (CNNs), for example, represent the flow of information from one layer to another as tensors, storing visual information densely across the visual topography, with different translation-invariant properties represented in different channels. CNNs have achieved state-of-the-art performance for some computer vision tasks, such as image recognition. These networks are also known to mimic many properties of biological systems, such that each can be attributed a ``Brain Score''~\citep{schrimpf_brain-score_2020}, yet this score takes no account of inference speed or energy consumption. Indeed, they do not take advantage of the dynamics inherent to the way we perceive our natural environment. In the vast majority of biological neural networks, on the other hand, information is represented as \emph{spikes}, prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that fired the spike~\citep{paugam-moisy_computing_2012}. Spiking neural networks (SNNs), known as the third generation of artificial neural networks, incorporate this temporal dimension into the way they perform their computations. They are of interest for computational neuroscience because they provide a better model of biological neuron. Some approaches have developed normative models of SNNs that aim to have applications in machine learning. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN that uses accurate timing of spikes attempts to determine the structure of the network to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been recently extended using a surrogate gradient, a method which is now widely used in attempts to transfer the performance of CNNs to SNNs~\citep{zenke_remarkable_2021}. However, this performance of SNNs still lags behind that of networks using an analog, firing-rate based representation. The question of the advantage of using spikes in machine learning and computer vision remains open.

In a recent review paper, we reported experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2023}. These spatio-temporal motifs which are present within the neural activity may form useful representations to perform computations for various cognitive tasks using the synchronicity of spikes reaching the soma of a neuron. Notably, \citet{abeles_role_1982} asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs. While the first possibility favors the firing-rate coding theory, the second one emphasizes the function of temporal precision in the neural code. Since then, numerous studies have demonstrated the emergence of synchronicity of activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient coding using spike latencies~\citep{perrinet_coding_2004, gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatio-temporal spiking motifs in biological spike trains. In neural models, the integration of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows for the efficient exploitation or detection of these spatio-temporal motifs embedded in the spike train. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a spiking motif defined by a subset of neurons with different, but precise, relative spiking delays. This delay between a pair of connected neurons is defined as the time between the emission of a spike on the afferent neuron's soma and that of its arrival on the efferent neuron's soma. Due to the variety of weights and delay within a population, representations using polychronous groups have mathematically a much higher information capacity than a firing rate-based approach.

The present paper proposes a real-life application extending a recent model of spiking neurons with heterogeneous synaptic delays~\cite{grimaldi_learning_2022}. Indeed, this previous model was trained to solve a simplified motion detection task on a synthetic event-based dataset generated by moving parameterized textures. This constituted a preliminary demonstration that formal neurons could exploit the precise timing of spikes to improve neural computations thanks to heterogeneous delays. Here, we extend these results to a more complex and natural framework. We first define the ecological cognitive task the model has to solve with the different datasets it will be tested on. Instead of the synthetic textures that have been used to generate motion-driven stimuli, we use natural images and random movements that mimic biological saccades. Then, we develop the Heterogeneous Delays Spiking Neural Network (HD-SNN) model thanks to theoretically defined first principles and derive a learning rule to adapt these heterogeneous synaptic delays. Using this detection task, we study the emergence of spatio-temporal spiking motifs when this single layer of spiking neurons is trained in a supervised manner. We will investigate the efficiency of the motion detection mechanism and in particular its resilience to synaptic weight pruning. Indeed, once trained, the amount of event-driven computations could be drastically reduced by pruning synapses, while maintaining top performance for the classification task. In this way, we will be able to show how such a model can provide an efficient solution to the trade-off between energy and accuracy.

>>>>

\keywords{\Keywords}
%
\maketitle
%
\section{Introduction}
%%%-----------------------------------------------------------------
\label{sec:intro}
%
The human brain has the remarkable capacity to react efficiently at any given time while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference with artificial neural networks is the representation that both use. Artificial convolutional neural networks (CNNs), for example, represent the flow of information from one layer to another as analog tensors, storing visual information densely across the visual topography, with different translation-invariant properties represented in different channels. CNNs have achieved state-of-the-art performance for some computer vision tasks, such as image recognition. These networks are also known to mimic many properties of biological systems, such that each can be assigned a ``brain score''~\citep{schrimpf_brain-score_2020}, but this score does not take into account inference speed or energy consumption. In fact, they do not take advantage of the dynamics inherent in the way we perceive our natural environment. In the vast majority of biological neural networks, on the other hand, information is represented as \emph{spikes}, prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that fired the spike~\citep{paugam-moisy_computing_2012}. Spiking neural networks (SNNs), known as the third generation of artificial neural networks, incorporate this temporal dimension into the way they perform their computations. They are of interest to computational neuroscience because they provide a better model of the biological neuron. One is the SpikeNet algorithm, which takes a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN that uses the exact timing of spikes tries to determine the structure of the network to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and was recently extended using a surrogate gradient, a method that is now widely used in attempts to transfer the performance of CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs still lags behind that of networks using an analog, firing rate-based representation. The question of the advantage of using spikes in machine learning and computer vision remains open.

In a recent review, we reported experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2023}. These spatiotemporal motifs present in neural activity may form useful representations to perform computations for various cognitive tasks using the synchrony of spikes reaching the soma of a neuron. In particular, \citet{abeles_role_1982} asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs. While the first possibility favors the firing rate coding theory, the second emphasizes the function of temporal precision in the neural code. Since then, numerous studies have demonstrated the emergence of synchronous activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient coding using spike latencies~\citep{perrinet_coding_2004, gollisch_rapid_2008}, or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatiotemporal spiking motifs in biological spike trains. In neural models, the integration of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows the efficient exploitation or detection of these spatio-temporal motifs embedded in the spike train. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a spiking motif defined by a subset of neurons with different, but precise, relative spiking delays. This delay between a pair of connected neurons is defined as the time between the emission of a spike at the soma of the afferent neuron and its arrival at the soma of the efferent neuron. Due to the variety of weights and delays within a population, representations using polychronous groups have a mathematically much higher information capacity than a firing rate-based approach.

The present paper proposes a real-world application that extends a recent model of spiking neurons with heterogeneous synaptic delays~\cite{grimaldi_learning_2022}. Indeed, this previous model was trained to solve a simplified motion detection task on a synthetic event-based dataset generated by moving parameterized textures. This was a first demonstration that formal neurons can exploit the precise timing of spikes to improve neural computation thanks to heterogeneous delays. Here, we extend these results to a more complex and natural setting. First, we define the ecological cognitive task that the model must solve with the different datasets on which it will be tested. Instead of the synthetic textures that have been used to generate motion-driven stimuli, we use natural images and random movements that mimic biological saccades. We then develop the Heterogeneous Delays Spiking Neural Network (HD-SNN) model thanks to theoretically defined first principles and derive a learning rule to adapt heterogeneous synaptic delays. Using this detection task, we study the emergence of spatio-temporal spiking motifs when this single layer of spiking neurons is trained in a supervised manner. We investigate the efficiency of the motion detection mechanism and, in particular, its resilience to synaptic weight pruning. Indeed, once trained, the amount of event-driven computation could be drastically reduced by removing weak synapses while maintaining peak performance for the classification task. In this way, we will be able to show how such a model can provide an efficient solution to the energy/accuracy tradeoff.
